---
title: "Homework 5: Binomial GLMs"
author: "Jamie Ash"
date: "due: 2021-09-29"
root.dir: "C:\\Users\\james\\Desktop\\OCN638\\HW4\\code"
output: 
  #html_document: 
  bookdown::html_document2:  
    self-contained: yes
    theme: paper #cerulean #cosmo #journal #readable
    toc: true
    smooth_scroll: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    fig_caption: yes
    code_folding: hide
    # bibliography: ["geothermal.bib", "r.bib"]
    # csl: apa-5th-edition.csl
    # link-citations: yes
---  

<style type="text/css">  
/* Note: CSS uses C-style commenting. */
h1.title{font-size:22px; text-align:center;}
h4.author{font-size:16px; text-align:center;}
h4.date{font-size:16px; text-align:center;}
body{ /* Normal  */ font-size: 13px}
td {  /* Table   */ font-size: 12px}
h1 { /* Header 1 */ font-size: 16px}
h2 { /* Header 2 */ font-size: 14px}
h3 { /* Header 3 */ font-size: 12px}
.math{ font-size: 10pt;}
.hi{ /* hanging indents */ 
    padding-left:22px; 
    text-indent:-22px;
}
blockquote {  
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 12px;
    border-left: 5px solid #eee;
}
code.r{ /* code */ 
       font-size: 12px;
}
pre{/*preformatted text*/ 
    font-size: 12px;
}
p.caption {/* figure captions */ 
    font-size: 1.0em;
    font-style: italic; 
} 
</style>

<style>
.vscroll-plot {
    width: 800px;
    height: 400px;
    overflow-y: scroll;
    overflow-x: hidden;
}
</style>

```{r setup, include=FALSE}
library(ggplot2)
library(bookdown)
library(ggplot2)
library(car)
library(effects)
library(ggeffects)
library(MASS)
library(GGally)
knitr::opts_chunk$set(echo = TRUE)
infile <- "..\\data\\"
```

Let’s consider some data on the prevalence of avian pox among birds on Hawaiʻi
island. In the attached paper the researchers investigated the prevalence of the 
disease at twelve sites in windward rainforests, over a range of altitudes, in four 
bird species – three endemic (ʻApapane, Hawaiʻi ʻAmakihi, ʻIʻiwi) and one 
introduced (Japanese White-eye). Individual birds were assayed for active pox 
infection (lesions), as well as chronic malaria infection (birds that survive malaria 
become chronically infected). ʻIʻiwi is highly susceptible to malaria, while 
ʻApapane and Hawaiʻi ʻAmakihi are moderately susceptible, and the White-eye is 
resistant.

# Display malaria in birds
The twelve sites have been coded as low/medium/high elevation. First, 
summarize the number of observations of each bird species at each elevation. This 
will be important context for what we can and cannot ask about avian pox 
patterns. Also, what is the prevalence of chronic malaria in the bird species at the 
difference elevations? Note that some birds do not have a known malaria status 
(code = 2). Only use birds with code = 0 or 1 to quantify the prevalence of malaria.
      
```{r}
birds  <- read.csv(paste(infile, "Pox_Data_revised.csv", 
                         sep = ""))
birds[birds$Malaria == 2, "Malaria"] <- NA

birds$x <- rep(1, nrow(birds))
elev  <- aggregate(birds$x, 
                   by = list(Species = birds$Species, Elev = birds$Elev),  
                   FUN = sum)
colnames(elev) <- c("Species", "Elev", "pop")

known <- subset(birds, !Malaria == 2)
sick_birds <- aggregate(known$Malaria, 
                   by = list(Species = known$Species, Elev = known$Elev),  
                   FUN = sum)
all_birds  <- aggregate(known$x, 
                   by = list(Species = known$Species, Elev = known$Elev),  
                   FUN = sum)

all_birds$mal_per <- round((sick_birds$x / all_birds$x) * 100, 0)
all_birds$mal_per <- paste(all_birds$mal_per, "%", sep = "")

mydata <- merge(elev, all_birds, by=c("Species", "Elev"), all = TRUE)
mydata[is.na(mydata$mal_per), ]$mal_per <- "NA"
```

**Answer**
It appears that the birds observed at higher elevations have a lower incident of chronic malaria. Bird species in mid elevation seem to have the highest incident of chronic malaria, although only the HAAM species has a reliable sample size. I find the malaria infection percentage using birds that had identifiable symptoms ie. `subset(birds, !Malaria == 2)`, but the population count depicts the total population of birds observed.

```{r, fig.cap = "Population of four bird species, Apapane (APAP), Hawaii Amakihi (HAAM), Iiwi (IIWI) and Japanese White-eye (JAWE) at three different elevations: low, mid, and high. The respective malaria infection rates are displayed above each bird's population bar."}
ggplot(mydata, aes(fill = Species, y = pop, x = Elev)) + 
  geom_bar(position = "dodge", stat = "identity") + 
  geom_text(aes(y = pop, label = mal_per), 
            vjust = -0.4, 
            position = position_dodge(width = 1)) +
  labs(title = "Prevailance of malaria in bird species at difference elevations") + 
  xlab("Elevation Type") +
  ylab("Population Count") +
  theme_bw()
```

# Pox birds without considering malaria
Create a model where the presence/absence of pox can differ between species 
and between elevations. Allow the effect of elevation to differ between species. 
Make an effects plot of the fitted model and perform likelihood ratio tests on the 
predictors. Provide a verbal explanation of what the model tells us. 
Look at summary() of the model, and note that there is an NA for one of the 
coefficients. Why do you think this is?

## Answer {.tabset}
Using a general linear model with a binomial family, and a type two Anova() likelihood test I found that there is a significant relationship between birds with pox and  bird species. I also found that there was a significant relationship between the effect of bird species on pox infection and elevation, as well as a significant direct relationship between pox infection and elevation. Because there are no Iiwi found at low elevations, the effects an NA is received for the Iwwi low coefficient. I ran a model (mod2) without the direct elevation interaction on pox infection and only Elev:Species, so that nice effects plots could be made. Lastly I combined active pox occurrences with old pox occurrences because I don't see why not, pox are pox. 
     
### Summary Statistics
```{r, out.width = "50%"}
birds  <- read.csv(paste(infile, "Pox_Data_revised.csv", 
                         sep = ""))

# Combining Active pox and Old Pox
birds$pox <- birds$Activepox + birds$Oldpox
birds[birds$pox == 2, "pox"] <- 1

# model to observe the Anova results
mod  = glm(pox ~ Species + Elev + Elev:Species, data = birds, family = binomial)
# model to plot the effects 
mod2 = glm(pox ~ Species + Elev:Species, data = birds, family = binomial)

summary(mod)
Anova(mod, type = 2)
```

### Effects Figures
```{r, out.width = "75%"}
plot(ggeffect(mod2, terms ="Elev"))
plot(ggeffect(mod2, terms ="Species"))
```

# Pox birds removing Malaria birds
Create a subset of the data that only includes birds with known malaria status (0 
or 1). Also, you’ll want to exclude ʻIʻiwi from this model (look at your answer to #1 
to see why). Create a model where pox prevalence varies by species, elevation, 
and malaria status, and where the effect of elevation and the effect of malaria status 
can differ between species. Make effects plots and perform likelihood ratio tests on 
the predictors. Provide your interpretation of what the results mean. 

```{r, out.width = "50%"}
birds  <- read.csv(paste(infile, "Pox_Data_revised.csv", 
                         sep = ""))

birds$pox <- birds$Activepox + birds$Oldpox
birds[birds$pox == 2, "pox"] <- 1

birds[birds$Malaria == 2, "Malaria"] <- NA
birds <- subset(birds, !is.na(Malaria))
birds <- subset(birds, !Species == "IIWI")

mod = glm(pox ~ Species + Elev + Elev:Species + Malaria + Malaria:Species, 
          data   = birds, 
          family = "binomial")

disp <- sum(residuals(mod, type = "pearson")^2)/(nrow(birds) - length(coef(mod)))
```

## Answer {.tabset}

Using a general linear model with `family = binomial` on a subset of the data without Iiwi or unknown malaria incidents, I found that there is a significant effect of Malaria, elevation, and species on the pox infection. I also found that there is a significant effect of elevation and malaria on the bird species effect on pox. Everything was significant. Also, I was confused by the wording "the effect of elevation and the effect of malaria status can differ between species" in the question. Is the model meant to be `pox ~ Species + Elev:Species + Malaria:Species` or `pox ~ Species + Elev + Elev:Species + Malaria + Malaria:Species` which is the same as `pox ~ Species * Elev * Malaria`. I went with the later. The dispersion parameter is `r round(disp, 2)` so I did not account for over dispersion. Lastly, I found that there is a positive near linear relationship between pox infection and malaria infection. 

### Summary Statistics 
```{r}
summary(mod)
Anova(mod)
```

### Effects Figures

```{r, out.width = "75%"}
plot(ggpredict(mod, terms ="Elev"))
plot(ggpredict(mod, terms ="Species"))
plot(ggpredict(mod, terms ="Malaria"))
```

# Do not pay attention to this

## Answer {.tabset}

I began doing question 2 and 3 in a similar way to the boar-TB example in lecture notes 7. I created a data frame with total population, and infected population so that I could do a glm() with `family = binomial`, and 'weights = population' where 'population' is the total bird observation per species/elevation. I decided the way I did it above was probably closer to the correct answer although the answers end up being very similar. I do the same thing as above where one test is run with Iiwi and one test is run without Iiwi or unknown malaria data. 

### Dataframe
```{r, out.width = "50%"}
birds  <- read.csv(paste(infile, "Pox_Data_revised.csv", 
                         sep = ""))

birds$pox <- birds$Activepox + birds$Oldpox
birds[birds$pox == 2, "pox"] <- 1

birds$x <- rep(1, nrow(birds))
all_birds <- aggregate(birds$x, 
                       by = list(Species = birds$Species, Elev = birds$Elev),  
                       FUN = sum)

pox_birds <- aggregate(birds$pox, 
                       by = list(Species = birds$Species, Elev = birds$Elev),  
                       FUN = sum)

all_birds$pox <- pox_birds$x
all_birds$per_pox <- pox_birds$x/all_birds$x
colnames(all_birds) <- c("Species", "Elev", "pop", "pox", "per_pox")

mod = glm(per_pox ~ Species + Elev:Species, 
          data    = all_birds, 
          family  = binomial, 
          weights = pop)

head(all_birds)
```

### Summary Statistics
```{r}
summary(mod)
Anova(mod, type = 2, test = "F")
```

### Effects Fiures
```{r}
plot(ggeffect(mod, terms ="Elev"))
plot(ggeffect(mod, terms ="Species"))
```


## Answer {.tabset}

### Dataframe
```{r, out.width = "75%"}
birds  <- read.csv(paste(infile, "Pox_Data_revised.csv", 
                         sep = ""))

birds[birds$Malaria == 2, "Malaria"] <- NA
birds <- subset(birds, !is.na(Malaria))
birds <- subset(birds, !Species == "IIWI")

birds$pox <- birds$Activepox + birds$Oldpox
birds[birds$pox == 2, "pox"] <- 1

birds$x <- rep(1, nrow(birds))
all_birds <- aggregate(birds$x, 
                       by = list(Species = birds$Species, Elev = birds$Elev),  
                       FUN = sum)

pox_birds <- aggregate(birds$pox, 
                       by = list(Species = birds$Species, Elev = birds$Elev),  
                       FUN = sum)

mal_birds <- aggregate(birds$Malaria, 
                       by = list(Species = birds$Species, Elev = birds$Elev),  
                       FUN = sum)

all_birds$mal <- mal_birds$x
all_birds$pox <- pox_birds$x
all_birds$per_mal <- mal_birds$x/all_birds$x
all_birds$per_pox <- pox_birds$x/all_birds$x
colnames(all_birds) <- c("Species", "Elev", "pop", "malaria","pox", "per_mal", 
                         "per_pox")

mod = glm(per_pox ~ Species + Elev + Elev:Species + per_mal + per_mal:Species, 
          data = all_birds, 
          family = "binomial", 
          weights = pop)
head(all_birds)
```

### Summary Statistics
```{r}
summary(mod)
Anova(mod, type = 2)
```

### Effects Figures
```{r}
plot(ggpredict(mod, terms ="Elev"))
plot(ggpredict(mod, terms ="Species"))
plot(ggpredict(mod, terms ="per_mal"))
```























