---
title: "Ash_A07"
author: "Jamie Ash"
date: "due: 2021-03-06"
output: 
  #html_document: 
  bookdown::html_document2:  
    self-contained: yes
    theme: cerulean #paper #cosmo #journal #readable
    toc: true
    smooth_scroll: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    fig_caption: yes
    code_folding: hide
    # bibliography: ["geothermal.bib", "r.bib"]
    # csl: apa-5th-edition.csl
    # link-citations: yes
---  

<style type="text/css">  
/* Note: CSS uses C-style commenting. */
h1.title{font-size:22px; text-align:center;}
h4.author{font-size:16px; text-align:center;}
h4.date{font-size:16px; text-align:center;}
body{ /* Normal  */ font-size: 13px}
td {  /* Table   */ font-size: 12px}
h1 { /* Header 1 */ font-size: 16px}
h2 { /* Header 2 */ font-size: 14px}
h3 { /* Header 3 */ font-size: 12px}
.math{ font-size: 10pt;}
.hi{ /* hanging indents */ 
    padding-left:22px; 
    text-indent:-22px;
}
blockquote {  
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 12px;
    border-left: 5px solid #eee;
}
code.r{ /* code */ 
       font-size: 12px;
}
pre{/*preformatted text*/ 
    font-size: 12px;
}
p.caption {/* figure captions */ 
    font-size: 1.0em;
    font-style: italic; 
} 
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bookdown)
```

# 
```{r Lab Cultures Mean}
```

#
```{r lm() and F-tests on Sex}
```

# 
```{r Model-fitted Group Means}
```

# 
```{r}
```

# 
```{r}
```

#
```{r}
```

#
```{r}
```













